\begin{aosachapter}{Moodle}{s:moodle}{Tim Hunt}

Moodle is a web application used in educational settings. While I will
try to give an overview of all aspects of how Moodle works, I will
focus on those areas where I think Moodle's design is particularly
interesting:

\begin{aosaenumerate}

\item the way the application is divided into plugins;

\item the permission system, which controls which users can perform
  which actions in different parts of the system;

\item the way output is generated, so that different themes (skins)
  can be used to give different appearances, and so that the interface
  can be localised; and

\item the database abstraction layer.

\end{aosaenumerate}

\begin{aosasect1}{What is Moodle?}

Moodle\footnote{\url{http://moodle.org/}} provides a place online
where students and teachers can come together to teach and learn. A
Moodle site is divided into Courses. A course will have certain users
enrolled in it in different roles, such as Student or Teacher. Each
course will comprise a number of resource and activities. A resource
might be a PDF file, a page of HTML within Moodle, or a link to
something else on the web. An activity might be a forum, a quiz or a
wiki. Within the course, these resources and activities will be
structured in some way. For example they may be grouped into logical
topics, or into weeks on a calendar.

% Moodle course.tiff
% FIXME: insert figure

Moodle can be used as a stand-alone application. If you wanted to
teach software architecture (for example) you could download Moodle to
your web hosting, install it, start creating courses, and wait for
students to come and self-register. Alternatively, if you are a large
institution, Moodle would be just one of the systems you run. You
would probably also have:

% \begin{aosafigure{../moodle/university-systems.png}{University Systems}{fig.moodle.university}
% FIXME: insert university systems diagram here (must convert from Dia format)

\begin{aosaitemize}

\item an authentication/identity provider (for example LDAP) to
  control user accounts across all your systems.


\item a student information system. That is, a database of all your
  students, which program of study they are one, and hence which
  courses they need to complete; and their transcript, that is a high
  level summary of the results of the courses they have
  completed. This would also deal with other administrative functions
  like tracking whether they have paid their fees.


\item a document repository (for example Alfresco). To store files,
  and track the workflow as users collaborate to create files.


\item an ePortfolio. This is a place where students can put assets and
  assemble them. This might be in order to build a CV (resume), or to
  assemble evidence that they have met the requirements of a
  work-based course.


\item reporting or analytics tools. To generate high level information
  about what is going on in your institution.

\end{aosaitemize}

Moodle focusses on providing a space online where people can engage in
teaching and learning, rather than any of the other systems that an
educational organisation might need. Moodle provides a basic
implementation of the other functionality, so that it can function as
a stand-alone system, or will integrate with other systems. The role
Moodle plays is normally called a virtual learning environment (VLE),
or learning or course management system (LMS, CMS or even LCMS).

Moodle is open source or free software (GPL). It is written in PHP. It
will run on most common web servers, on common platforms. It requires
a database, and will work with MySQL, PostgreSQL, MS SQL Server or
Oracle.

The Moodle project was started by Martin Dougiamas in 1999, while he
was working at Curtin University, Australia. Version 1.0 was released
in 2002, at which time PHP4.2 and MySQL 3.23 were the technologies
available. That limited the kind of architecture that was possible
initially, but much has changed since then. The current release is the
Moodle 2.2.x series.

\end{aosasect1}

\begin{aosasect1}{How Moodle Works}

A Moodle installation comprises three parts:

\begin{aosaenumerate}

\item The code, typically in a folder like \code{/var/www/moodle} or
  \code{{\textasciitilde}/htdocs/moodle}. This should not be writable
  by the web server.

\item The database, managed by one of the supported RDMSs. In fact,
  Moodle adds a prefix to all the table names, so it can share a
  database with other applications, if desired.

\item The \code{moodledata} folder. This is a folder where Moodle
  stores uploaded and generated files, and so needs to be writable by
  the web server. For security reasons, the should be outside the web
  root.

\end{aosaenumerate}

These can all be on a single server. Alternatively, in a load-balanced
set-up, there will be multiple copies of the code on each web-server,
but just one shared copy of the database and moodledata, probably on
other servers.

The configuration information about these three parts is stored in a
file called \code{config.php} in the root of the \code{moodle} folder
when Moodle is installed.

\begin{aosasect2}{Request Dispatching}

Moodle is a web applications, so users interact with it using their
web browser. From Moodle's point of view that means responding to HTTP
requests. An important aspect of Moodle's design is, therefore, the
URL namespace, and how URLs get dispatched to different scripts.

Moodle uses the standard PHP approach to this. To view the main page
for a course, the URL would be \url{.../course/view.php?id=123}, where
123 is the unique id of the course in the database. To view a forum
discussion, the URL would be something like
\url{.../mod/forum/discuss.php?id=456789}. That is, these particular
scripts, \url{course/view.php} or \url{mod/forum/discuss.php} would
handle these requests. This is simple. To understand how Moodle
handles a particular request, you look at the URL and start reading
code there.

The alternative approach one could take is to have a single entry
point \url{.../index.php/\emph{extra}}, where \emph{extra} is
information needed to make the request unique. The single script
\code{index.php} would then dispatch the requests in some way. This
approach adds a layer of indirection, which is something software
developers always like to do. The lack of this layer of indirection
does not seem to hurt Moodle. (In the rare cases where you really want
to rewrite the URL namespace, this can be done elsewhere in the
software stack, for example by using \code{mod\_rewrite}.)

\end{aosasect2}

\begin{aosasect2}{Plugins}

Like many successful open source projects, Moodle is built out of many
plugins, working together with the core of the system. This is a good
approach because at allows people to change an enhance Moodle in
defined ways. An important advantage of an open source system is that
you can tailor it to your particular needs. Making extensive
customisations to the code can, however, lead to big problems when the
time comes to upgrade, even when using a good version control
system. By allowing as many customisations and new features as
possible to be implemented as self-contained plugins that interact
with Moodle core through a defined API, it easier for people to
customise Moodle to their needs, and to share customisations, while
still being able to upgrade the core Moodle system.

There are various ways a system can be built as a core surrounded by
plugins. Moodle has a relatively fat core, and the plugins are
strongly-typed. When I say a fat core, I mean that there is a lot of
functionality in the core. This contrasts to the kind of architecture
where just about everything, except for a small plugin-loader stub, is
a plugin.

When I say plugins are strongly type, I mean that depending on which
type of functionality you want to implement, you have to write a
different type of plugin, and implement a different API. For example,
an new Activity module plugin would be very different from a new
Authentication plugin or a new Question type. At the last count there
are about 35 different types of plugin. This contrasts with the kind
of architecture where all plugins use basically the same API and then,
perhaps, subscript the the subset of hooks or events they are
interested in.

Generally, the trend in Moodle has been to try to shrink the core, by
moving more functionality into plugins. This is only somewhat
successful, however, because an increasing feature-set tends to expand
the core. The other trend has been to try to standardise the different
types of plugin as much as possible, so that in areas of common
functionality, like install and upgrade, all types of plugins work the
same way.

A plugin is Moodle takes the form of a folder containing files. The
plugin has a type and a name, which together make up the
'Frankenstyle' component name of the plugin. (the word Frankenstyle
arouse out of an argument in the developers' Jabber channel, but
everyone liked it and it stuck.) The plugin type and name determine
the path to the plugin folder. The plugin type gives a prefix, and
then the foldername is the plugin name. Here are some examples.

\begin{tabular}{llll}
Plugin type & Plugin name & Frankenstyle & Folder \\
mod (Activity module) & forum & mod\_forum & mod/forum \\
mod (Activity module) & quiz & mod\_quiz & mod/quiz \\
block (Side-lock) & navigation & block\_navigation & blocks/navigation \\
qtype (Question type) & shortanswer & qtype\_shortanswer & question/type/shortanswer \\
quiz (Quiz report) & statistics & quiz\_statistics & mod/quiz/report/statistics \\
\end{tabular}

The last example shows that each activity module is allowed to declare
sub-plugin types. At the moment, of all the different types of plugin,
only activity modules are allowed to do this.

\end{aosasect2}

\begin{aosasect2}{An Example Plugin}

I will explain a lot of details of the Moodle architecture by
considering a specific example plugin. As is traditional, I have
chosen to implement a plugin that displays ``Hello world''.

This plugin does not really fit naturally into any of the standard
Moodle plugin types. It is just a script, with no connection to
anything else, so I will choose to implement it as a \emph{local}
plugin. This is a catch-all plugin type for miscellaneous
functionality that does not fit anywhere better. I will name my plugin
greet, to give a Frankensyle name of \code{local\_greet}, and a folder
path of \code{local/greet}. The plugin code can be downloaded from
\url{https://github.com/timhunt/moodle-local\_greet}.

Each plugin must contain a file called \code{version.php} which
defines some basic meta-data about the plugin. This is used by the
Moodle's plugin installer system to install and upgrade the plugin.
Ours is:

\begin{verbatim}
<?php
$plugin->component    = 'local_greet';
$plugin->version      = 2011102900;
$plugin->requires     = 2011102700;
$plugin->maturity     = MATURITY_STABLE;
\end{verbatim}

It may seem redundant to include the component name, since this can be
deduced from the path, but the installer uses this to verify that the
plugin has been installed in the right place. The version field is the
version of this plugin. Maturity is ALPHA/BETA/RC/STABLE. Requires is
the minimum version of Moodle that this plugin is compatible with. If
necessary, one can also document other plugins that this one depends
on.

Now here is the main scrip for this simple plugin, which we put
in \code{local/greet/index.php}:

\begin{verbatim}
<?php
require_once(dirname(__FILE__) . '/../../config.php');        // 1

require_login();                                              // 2
$context = context_system::instance();                        // 3
require_capability('local/greet:begreeted', $context);        // 4

$name = optional_param('name', '', PARAM_TEXT);               // 5
if (!$name) {
    $name = fullname($USER);                                  // 6
}

add_to_log(SITEID, 'local_greet', 'begreeted',
        'local/greet/index.php?name=' . urlencode($name));    // 7

$PAGE->set_context($context);                                 // 8
$PAGE->set_url(new moodle_url('/local/greet/index.php'),
        array('name' => $name));                              // 9
$PAGE->set_title(get_string('welcome', 'local_greet'));       // 10

echo $OUTPUT->header();                                       // 11
echo $OUTPUT->box(get_string('greet', 'local_greet',
        format_string($name)));                               // 12
echo $OUTPUT->footer();                                       // 13
\end{verbatim}

\begin{aosasect3}{Line 1: Bootstrapping Moodle}

The single line of this script that does the most work is the first. I
said above that config.php contains the details Moodle needs to
connect to the database and find the moodledata folder. It ends,
however, with the line \code{require\_once('lib/setup.php')}. This:

\begin{aosaenumerate}

\item loads all the standard Moodle libraries using \code{require\_once};

\item starts the session handling;

\item connects to the database; and

\item sets up a number of global variables, which we shall meet later.

\end{aosaenumerate}

\end{aosasect3}

\begin{aosasect3}{Line 2: Checking the User Is Logged In}

This line causes Moodle to check that the current user is logged in,
using whatever authentication plugin the administrator has
configured. If not, the user will be redirected to log in, and this
function will never return.

If we were writing a script that was more integrated into Moodle, we
would pass more arguments here, to say which course or activity this
page is part of, and then \code{require\_login} would also verify that
the user is enrolled in, or otherwise allowed to access this course,
and is allowed to see this activity. If not, an appropriate error
would be displayed.

\end{aosasect3}

\begin{aosasect3}{Line 3: Getting the Context}

The next two lines of code show how to check that the user has
permission to do something. As you can see, from the developer's point
of view, the API is very simple. Behind the scenes, however, there is
a sophisticated access system which gives the administrator great
flexibility to control who can do what.

In Moodle, users can have different permissions in different
places. For example, a user might be a Teacher in one course, and a
Student in another, and so have different permissions in each
place. These places are called contexts. Contexts in Moodle form a
hierarchy rather like a folder hierarchy in a file-system. At the top
level is the system context (and, since our script is not very well
integrated into Moodle, we use that context).

Within the system context are a number of contexts for the different
categories that have been created to organise courses. These can be
nested, with one category containing other categories. Category
contexts can also contain Course contexts. Finally, each activity in a
course will have its own Module context.

\begin{verbatim}
System
   Humanities faculty
      History department
         History 101
            ...
         Napoleon 202
            Social forum
            Resource: Napoleon biography
            Forum: was Napoleon right?
            Quiz: Napoleon facts
            ...
      Art department
         Art 101
         ...
      Introduction to the humanities
   Science faculty
      ...
   ...
\end{verbatim}

\end{aosasect3}

\begin{aosasect3}{Line 4: Checking the User Has Permission to Use This Script}

Now we have got the context---the relevant area of Moodle to check the
permission in---we now need to decide which permission to check. Each
particular area of functionality that a user may or may not have is
called a Capability. Checking a capability provides more fine-grained
access control than the basic checks performed by
\code{require\_login}. For our simple example plugin, we have one
capability: \code{local/greet:begreeted}.

The check is done using the \code{require\_capability} function, which
takes the capability name and the context. Like other
\code{require\_\ldots} functions, it will not return if the user does
not have the capability. It will display an error instead. In other
places we would use the non-fatal \code{has\_capability} function,
which returns a boolean. We might use that, for example, to determine
whether to display a link to this script from another page.

How does the administrator configure which user has which permission?
Here is the calculation that \code{has\_capability} performs (at least
conceptually):

\begin{aosaenumerate}

\item Start from the current Context.

\item Get a list of the Roles that the user has in this Context.

\item Then work out what the Permission is for each Role in this
  Context.

\item Aggregate those permissions to get a final answer.

\end{aosaenumerate}

As the example shows, a plugin can define new capabilities, relating
to the particular functionality it provides. This is done using a file
called access.php. This is stored inside a db sub-folder in the
plugin, which is where all the information required to install or
upgrade the plugin lives. Here is the \code{access.php} file for our
plugin, which we put in \code{local/greet/db/access.php}:

\begin{verbatim}
<?php
$capabilities = array('local/greet:begreeted' => array(
    'captype' => 'read',
    'contextlevel' => CONTEXT_SYSTEM,
    'archetypes' => array('guest' => CAP_ALLOW, 'user' => CAP_ALLOW)
));
\end{verbatim}

This gives some metadata about each capability which are used when
constructing the permissions management user-interface. We also give
default permissions for common types of role.

\end{aosasect3}

\begin{aosasect3}{Roles}

The next part of the Moodle permissions system is roles. A role is
really just a named set of permissions. When you are logged into
Moodle, you will have the 'Authenticated user' role in the system
context, and since the system context is the root of the hierarchy,
that role will apply everywhere.

Within a particular course, you may be a student, and that role
assignment will apply in the course context, and all the module
contexts within it. In another course, however, you may have a
different role. For example, Mr Gradgrind may be teacher in the
'Facts, Facts, Facts' course, but a Student in the professional
development course 'Facts aren't everything'. Finally, a user might be
given the Moderator role in one particular forum (Module context).

\end{aosasect3}

\begin{aosasect3}{Permissions}

A role defines a permission for each capability. For example, we will
probably say that the teacher role ALLOWS moodle/course:manage, but
the student role does not. However, both Student and Teacher will
allow \code{mod/forum:startdiscussion}.

The roles are normally defined globally, but they can be re-defined in
each context. For example, if we want to make one particular wiki
activity read-only to students of the course, then we would override
the permissions for the Student role in that wiki (Module) context, to
set the permission for \code{mod/wiki:edit} to \code{PREVENT}.

There are four Permissions:

\begin{aosaitemize}

\item NOT SET / INHERIT (default)

\item ALLOW

\item PREVENT

\item PROHIBIT (example coming up soon)

\end{aosaitemize}

In a given context, a role will have one of these four permissions for
each capability. One different between PROHIBIT and PREVENT is that a
PROHIBIT cannot be overridden in sub-contexts.

\end{aosasect3}

\begin{aosasect3}{Permission Aggregation}

Finally we aggregate the permissions for all the roles the user has in
this context.

\begin{aosaitemize}

\item If any role gives the permission PROHIBIT for this capability,
  return false.

\item Otherwise, if any role gives ALLOW for this capability, return
  true.

\item Otherwise return false.

\end{aosaitemize}

A use case for PROHIBIT is this: Suppose a user has been making
abusive posts in a number of forums, and we want to stop them
immediately. We can create a Naughty user role, which sets
\code{mod/forum:post} and other such capabilities to PROHIBIT. We can
then assign this role to the abusive user in the system context. That
way, we can be sure that the user will not be able to post any more in
any forum. (One would then talk to the student, and having reached a
satisfactory outcome, removing that role assignment so that they my
use the system again.)

So, Moodle's permissions system gives administrators a huge amount of
flexibility. They can define whichever roles they like with different
permissions for each capability; they can alter the role definitions
in sub-contexts; and then they can assign different roles to users in
different contexts.

Now, back to our example script{\ldots}

\end{aosasect3}

\begin{aosasect3}{Line 5: Get Data From the Request}

Something that every web application has to do is to get data from the
request (GET or POST variables) without being susceptible to SQL
injection or cross-site scripting attacks. Moodle provides two ways to
do this.

The simple method is the one shown here. To get a single variable, we
give the parameter name (here 'name') a default value, and the
expected type. The expected type is used to clean the input of all
unexpected characters. There are numerous types like
\code{PARAM\_INT}, \code{PARAM\_ALPHANUM}, \code{PARAM\_EMAIL} and so
on.  There is also a similar \code{required\_param} function, which
gives a fatal error if an expected parameter is not found.

The other way Moodle has to get data from the request is a
fully-fledged forms library. This is a wrapper around the HTML
QuickForm library from PEAR. (For non-PHP programmers, PEAR is PHP's
equivalent of CPAN.) This seemed like a good choice when we selected
it, but is now no longer maintained, so at some time in the future we
will have to tackle moving to a new forms library, which many of us
look forwards to, because QuickForm has several irritating design
issues. For now, however, it is adequate. Forms can be defined as a
collection of fields of various types (e.g. text box, select
drop-down, date-selector) with client- and server-side validation
(including use of the same \code{PARAM\_\ldots} types).

\end{aosasect3}

\begin{aosasect3}{Line 6: Global Variables}

Here we see the first of the global variables Moodle
provides. \code{\$USER} makes accessible the information about the
user accessing this script. Other globals include:

\begin{aosaitemize}

\item \code{\$CFG}: holds the commonly used configuration settings.

\item \code{\$DB}: the database connection.

\item \code{\$SESSION}: a wrapper around the PHP session.

\item \code{\$COURSE}: the course we are within.

\item And several others, some of which we will encounter later.

\end{aosaitemize}

You may have read the words ``global variable'' with horror. Note,
however, that PHP processes a single request at a time. Therefore
these variables are not as global as all that. In fact, PHP global
variables can be seen as an implementation of the Thread-scoped
registry pattern (in the sense of Martin Fowler, \emph{Patterns of
  Enterprise Application Architecture}) and this is the way in which
Moodle uses them. It is very convenient in that it makes commonly used
objects available thought the request without requiring them to be
passed to every function and method. It is only infrequently abused.

\end{aosasect3}

\begin{aosasect3}{Line 6 Again: Nothing Is Simple}

This line also serves to make a point about the problem
domain. Nothing is ever simple. Here, we just want to display a user's
name, but we can't simply echo:

\begin{verbatim}
$USER->firstname, ' ', $USER->lastname
\end{verbatim}

The school may have policies about showing either of those parts, and
different cultures have different conventions for which order to show
names. Therefore, there are several configurations settings, and we
have a function to assemble the full name according to the rules.

Similarly with dates. Different users may be in different
timezones. Moodle stores all dates as Unix timestamps, which are
integers, and so work simply cross-DB. There is then a
\code{userdate()} function to display the timestamp to the user using
the appropriate timezone and locale settings.

\end{aosasect3}

\begin{aosasect3}{Line 7: Logging}

All significant actions in Moodle are logged. Logs are written to a
table the database. This is a trade-off. It makes sophisticated
analysis quite easy, and indeed various reports base on the logs are
included with Moodle. On a large and busy site, however, it is a
performance problem. The log table gets huge, which makes backing up
the database more difficult, and makes queries on the log table
slow. There can also be write contention on the log table. These
problems can be mitigated in various ways, for example by batching
writes, or archiving or deleting old records to remove them from the
main database.

\end{aosasect3}

\end{aosasect2}

\end{aosasect1}

\begin{aosasect1}{Generating Output}

Output is mainly handled via two global objects.

\begin{aosasect2}{Line 8: The \code{\$PAGE} Global}

\code{\$PAGE} stores information about the page we are going to
output, that will be needed when the time comes to generate the
HTML. In this script, we need to explicitly specify the current
context. (In other situations, this might have been set automatically
by \code{require\_login}.) We also need to set the URL for this
page. This may seem redundant. The rationale for requiring it is that
you might get to a particular page using any number of different URLs,
but the URL passed to \code{set\_url} should be the canonical URL for
this page---a good permalink, if you like. We also specify the page
title, which will end up in the head element of the HTML.

\end{aosasect2}

\begin{aosasect2}{Line 9: Moodle URL}

I just wanted to flag this nice little helper class we have. Note, as
a sign of the age of Moodle's code-base, that I was not able to use it
in the call to the \code{add\_to\_log} function.

\end{aosasect2}

\begin{aosasect2}{Line 10: Internationalisation}

Moodle uses its own system to allow the interface to be translated
into any language. There may now be good PHP internationalisation
libraries, but in 2002 when it was first
implemented\footnote{\url{(http://git.moodle.org/gw?p=moodle.git;a=commitdiff;h=1180c6dc56dba2b3938119584044c69c8de62725}}
there was not one available that was adequate. The system is based
around the \code{get\_string} function. Strings are identified by a
key and the plugin Frankenstyle name. As can be seen on line 12, it is
possible to interpolate values into the string. (Multiple values are
handled using PHP arrays or objects.)

The strings are looked up in language files that are just plain PHP
arrays. Here is the language file for our plugin, which we put in
\code{local/greet/lang/en/local\_greet.php}:

\begin{verbatim}
<?php
$string['greet:begreeted'] = 'Be greeted by the hello world example';
$string['welcome'] = 'Welcome';
$string['greet'] = 'Hello, {$a}!';
$string['pluginname'] = 'Hello world example';
\end{verbatim}

Note that, as well as the two string used in our script, we also have
to define a string to give a name to our capability, and we have to
define a string \code{'pluginname'} which is the name of our plugin as
it appears in the user interface.

The different languages are identified by the two-letter country code
(\code{en} here). Languages packs may derive from other language
packs. For example the \code{fr\_ca} (French canadian) language pack
declares \code{fr} (French) as the parent language, and thus only has
to define those strings that differ from the French. Since Moodle
originated in Australia, \code{en} means British English, and
\code{en\_us} (American English) is derived from it.

Again, the simple \code{get\_string} API for plugin developers hides a
lot of complexity, including working out the current language (which
may depend on the current user's preferences, or the settings for the
particular course they are currently in), and then searching through
all the language packs and parent language packs to find the string.

Producing the language pack files, and co-ordinating the translation
effort is managed at \url{http://lang.moodle.org/}, which is Moodle
with a custom plugin\footnote{\code{local\_amos},
\url{http://docs.moodle.org/22/en/AMOS}.}. It uses both Git and the
  database as a back-end to store the language files with full version
  history.

\end{aosasect2}

\begin{aosasect2}{Line 11: Starting Output}

This is another innocuous-looking line that does much more than it
seems. The point is that before any output can be done, the applicable
theme (skin) to use must be worked out. The theme may depend on a
combination which context we are in and the user's
preferences. \code{\$PAGE->context} was, however, only set on line
8. In order to solve the problem, some PHP magic is used so that the
first time a method on \code{\$OUTPUT} is called, it re-initialises
itself based on the information in \code{\$PAGE}, and only then
generates the output.

Another thing that needs to be done is that every page in Moodle may
contain \emph{blocks}. These are extra configurable bits of content
that are included in the page. (They are a type of plugin.) Again, the
exact collation of blocks to display depends in a flexible way (that
the administrator can control) on the page context and some other
aspects of the page identity. Therefore, another part of preparing for
output is a call to
\code{\$PAGE-{\textgreater}blocks-{\textgreater}load\_blocks()}.

Once all the necessary information has been worked out, the theme
plugin (that controls the overall look of the page) is called to
generate the overall page layout, including whatever standard header
and footer is desired. This call is also responsible for adding the
output from the blocks at the appropriate place in the HTML. In the
middle of the layout there will be a div where the specific content
for this page goes. The HTML of this layout is generated, and then
split in half after the start of the main content div. The first half
is returned, and the rest is stored to be returned by
\code{\$OUTPUT-{\textgreater}footer()}.

\end{aosasect2}

\begin{aosasect2}{Line 12: Outputting the Body of the Page}
 
This line outputs the body of the page. Here we simply display the
greeting in a box. The greeting is again a localised string, this time
with a value substituted into a placeholder. The core renderer
\code{\$OUTPUT} provides many convenience methods like \code{box()} to
describe the required output in quite high-level terms. Different
themes can control what HTML is actually output to make the box.

The content that originally came from the user (\code{\$name}) is
output though the \code{format\_string} function. This is the other
part of providing XSS protection. It also enables the user of text
filters (another plugin type). An example filter would be the LaTeX
filter, which replaces input like \code{\$x + 1\$} with an image of
the equation. I will mention, but not explain, that there are actually
three different functions (\code{s()}, \code{format\_string()}, and
\code{format\_text()}) depending on the particular type of content
being output.

\end{aosasect2}

\begin{aosasect2}{Line 13: Finishing Output}

Finally, we have to output the footer of the page. This example does
not show it, but Moodle tracks all the JavaScript that is required by
the page, and outputs all the necessary script tags in the
footer. This is standard good practice. It allows users to see the
page without waiting for all the JavaScript to load. A developer would
include JavaScript using API calls like
\code{\$PAGE-{\textgreater}requires-{\textgreater}js('/local/greet/cooleffect.js')}.

\begin{aosasect3}{Should We Be Mixing Logic and Output?}

Obviously, putting the output code directly in index.php, even if at a
high level of abstraction, limits the flexibility that themes have to
control the output. This is another sign of the age of the Moodle
code-base. The \code{\$OUTPUT} global was introduced in 2010 as a
stepping stone on the way from the old code, where the output and
controller code was in the same file, to a design where all the view
code was properly separated. This also explains the rather ugly way
that the entire page layout is generated, then split in half, so that
any output from the script itself can be placed between the header and
the footer. Once the view code has been separated out of the script,
into what Moodle calls a renderer, the theme can then choose to
completely (or partially) override the view code for a given script.

With a small refactoring, we can go the final step and move all the
output code out of our index.php and into a renderer. We could change
the end of \code{index.php} (lines 11--13) to:

\begin{verbatim}
$output = $PAGE->get_renderer('local_greet');
echo $output->greeting_page($name);
\end{verbatim}

and then make a new file \code{local/greet/renderer.php}:

\begin{verbatim}
<?php
class local_greet_renderer extends plugin_renderer_base {
    public function greeting_page($name) {
        $output = '';
        $output .= $this->header();
        $output .= $this->box(get_string('greet', 'local_greet', $name));
        $output .= $this->footer();
        return $output;
    }
}
\end{verbatim}

If the theme wished to completely change this output, it would define
a subclass of this renderer that overrides the \code{greeting\_page}
method. \code{\$PAGE-{\textgreater}get\_renderer()} determines the
appropriate renderer class to instantiate depending on the current
theme. Thus, the output (view) code is fully separated from the
controller code in index.php, and we have been able to refactor from
typical legacy Moodle code to a clean MVC architecture.

\end{aosasect3}

\end{aosasect2}

\end{aosasect1}

\begin{aosasect1}{Database Abstraction}

TODO

\end{aosasect1}

\begin{aosasect1}{What I Have Not Covered}

I hope I have given you a good overview of how Moodle works. Due to
lack of space I have had to omit several interesting topics, including
how authentication, enrolment and grade plugins allow Moodle to
interoperate with student information systems, and the interesting
content-addressed way that Moodle stores uploaded files. Details of
these, and other aspects of Moodle's design can be found in the
developer documentation: http://docs.moodle.org/dev/.

\end{aosasect1}

\begin{aosasect1}{Lessons Learned}

TODO turn these bullets into proper writing.

\begin{aosaitemize}

\item Nothing is as simple as it seems.

\item Still, start with the simple solution, and only move to
  something more complex when the need arises. You \emph{can} change
  the architecture later---and by then you will have a much better
  idea what is needed. Later, the technology may have moved on, making
  it easier to implement the complex thing.

\item The most profound thing I have learned from Moodle, which I have
  not yet alluded to, is that writing software is a learning
  process. Working on educational software is a good way to realise
  this, particularly Moodle which embraces a Social Constructivist
  pedagogy. I strongly recommend the article
  \url{http://docs.moodle.org/en/Pedagogy}. Don't think of building
  your software as a problem to solve. Rather, think of it as
  something you have to learn. Something you can only learn by doing
  it, and by iterating with others, particularly your users.

\end{aosaitemize}

\end{aosasect1}

\end{aosachapter}
